The Calculus of Linear Constructions

### 2021-10-19 writing/lics22/lics22.tex
lics22

\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{7pt}
\setlength{\grammarindent}{7em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\sort}{\textbf{sort}}
\newcommand{\type}{\textbf{type}}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{\ $\text{let }#1\text{ := }#2\text{ in }#3$\ }
\newcommand{\ind}{\ $\text{ind}$\ }
\newcommand{\iter}{\ $\text{iter}$\ }
\newcommand{\new}{\text{new}}
\newcommand{\free}{\text{free}}
\newcommand{\get}{\text{get}}
\newcommand{\set}{\text{set}}
\newcommand{\utype}{:_{\scriptscriptstyle U}}
\newcommand{\ltype}{:_{\scriptscriptstyle L}}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}
\newcommand{\erase}[1]{\llbracket #1 \rrbracket}

@@

  \section{Extensions}
  Readers familiar with the literature on linear logic or linear type theory will have noticed at this point that CLC does not possess the famous $\otimes$ connective. Though it is entirely possible to encode $\otimes$ in core CLC through a linear Church encoding, proofs and programs written in this style will be incredibly difficult to use.

@@

  \begin{figure}[h]
    \vspace{-0.8em}
    \caption{Select Formation Rules}
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash \mathbf{1} : U_0 }
      \rname{$1$-Form}

@@

  \begin{figure}[h]
    \vspace{-0.8em}
    \caption{Select Introduction Rules}
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash () : \mathbf{1} }
      \rname{$1$-Intro}

@@

      \inferrule
      { \Gamma_1 \vdash m : A \\
        \Gamma_2 \vdash n : B \\ 
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\\\
        \Gamma \vdash A \otimes B : L_k }
      { \Gamma \vdash (m, n) : A \otimes B }
      \rname{$\otimes$-Intro}

      \inferrule
      { \pure{\Gamma_1} \\
        \Gamma_1 \vdash m : A \\
        \Gamma_2 \vdash n : B[m/x] \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\
        \Gamma \vdash F x:A.B : L_k }
      { \Gamma \vdash (m, n) : F x:A.B }
      \rname{$F$-Intro}
    \end{mathpar}
    \vspace{-2em}
    \label{introduction}
    \Description{}
  \end{figure}

  \begin{figure}[h]
    \vspace{-0.8em}
    \caption{Select Elimination Rules}
    \begin{mathpar}
      \inferrule
      { \Gamma_1 \vdash m : \mathbf{1} \\ 
        \Gamma_2 \vdash n : A \\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} }
      { \Gamma \vdash \letin{()}{m}{n} : A }
      \rname{$1$-Elim}

@@

      \inferrule
      { \Gamma_1 \vdash m : A \otimes B \\ 
        \Gamma_2, x \ltype A, y \ltype B \vdash n : C \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} }
      { \Gamma \vdash \letin{$(x, y)$}{$m$}{$n$} \vdash C }
      \rname{$\otimes$-Elim}

      \inferrule
      { \Gamma_1 \vdash m : F x:A.B \\
        \Gamma_2, x \utype A, y \ltype B \vdash n : C \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} }
      { \Gamma \vdash \letin{$(x, y)$}{$m$}{$n$} \vdash C }
      \rname{$F$-Elim}
    \end{mathpar}
    \vspace{-0.8em}
    \label{elimination}
    \Description{}
  \end{figure}

@@

  \subsection{Stateful Reasoning}
  We first postulate five state axioms. These axioms could be viewed as an interface to a trusted memory allocator.
  \begin{definition}
    State axioms.
    \begin{itemize}
      \item $l \mapsto_A m : \mathbb{N} \rightarrow (A : U_k) \rightarrow A \rightarrow L_k$ \\
      The $\mapsto$ is a linear type constructor. It fulfills a role similar to L3's \cite{l3} memory access capability, or Separation Logic's \cite{reynolds02} points-to assertion. Intuitively, it is a proof that a term $m$ of non-linear type $A$ is currently stored at address $l$.
      \item $\new : (A : U_k) \rightarrow (m : A) \rightarrow F l:\mathbb{N}.(l \mapsto_A m)$ \\
      $\new$ is a function that allocates a new memory store for term $m$ of non-linear type $A$. It results in an address $l$ paired with an assertion $l \mapsto_A m$.
      \item $\free : (A : U_k) \rightarrow F l:\mathbb{N}.F m:A.(l \mapsto_A m) \rightarrow \mathbf{1}$ \\
      $\free$ is a function that de-allocates the memory store at location $l$ which is currently in use. This can be seen in its input $F l:\mathbb{N}.F m:A.(l \mapsto_A m)$, where the $(l \mapsto_A m)$ component is a proof that location $l$ is currently allocated.
      \item $\get : (A : U_k) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \rightarrow F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$ \\
      $\get$ is a function that retrieves the term stored at address $l$ when given a proof $(l \mapsto_A m)$ that there is indeed a term $m$ currently stored there. The type of $\get$'s output is the rather complicated $F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$. This not only returns the term stored at $l$, but also a proof that the returned term is exactly the same as what $(l \mapsto_A m)$ claimed. The proof $(l \mapsto_A m)$ is returned unchanged as well.
      \item $\set : (A\ B : U_k) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \multimap (n : B) \multimap (l \mapsto_B n)$ \\
      $\set$ has the most complicated type of all the axioms presented here. Intuitively, if address $l$ is allocated and storing some term $m$ of type $A$, $\set$ may put term $n$ of type $B$ into the the store at address $l$, overwriting the original stored term. This is the so called "strong update". The consumption of $(l \mapsto_A m)$ and return of $(l \mapsto_B n)$ encode this update process.
    \end{itemize}
  \end{definition}

  \section{Future Work}
  \section{Conclusion}

>>>>

\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{7pt}
\setlength{\grammarindent}{7em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\sort}{\textbf{sort}}
\newcommand{\type}{\textbf{type}}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{\ $\text{let }#1\text{ := }#2\text{ in }#3$\ }
\newcommand{\ind}{\ $\text{ind}$\ }
\newcommand{\iter}{\ $\text{iter}$\ }
\newcommand{\unit}{\text{unit}}
\newcommand{\new}{\textbf{new}}
\newcommand{\free}{\textbf{free}}
\newcommand{\get}{\textbf{get}}
\newcommand{\set}{\textbf{set}}
\newcommand{\utype}{:_{\scriptscriptstyle U}}
\newcommand{\ltype}{:_{\scriptscriptstyle L}}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}
\newcommand{\erase}[1]{\llbracket #1 \rrbracket}

@@

  \section{Extensions}
  Readers familiar with the literature on linear logic or linear type theory will have noticed at this point that CLC does not possess the famous $\otimes$ connective. Though it is entirely possible to encode $\otimes$ in core CLC through a linear Church encoding, but proofs and programs written in this style will be incredibly difficult to maintain.

@@

  \begin{figure}[h]
    \vspace{-0.8em}
    \caption{Select Formation Rules}
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash \unit : U_0 }
      \rname{Unit-Form}

@@

  \begin{figure}[h]
    \vspace{-0.8em}
    \caption{Select Introduction Rules}
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash () : \unit }
      \rname{Unit-Intro}

@@

      \inferrule
      { \Gamma_1 \vdash m : A \\
        \Gamma_2 \vdash n : B \\ 
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\\\
        \Gamma \vdash A \otimes B : L_k }
      { \Gamma \vdash [m, n] : A \otimes B }
      \rname{$\otimes$-Intro}

      \inferrule
      { \pure{\Gamma_1} \\
        \Gamma_1 \vdash m : A \\
        \Gamma_2 \vdash n : B[m/x] \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\
        \Gamma \vdash F x:A.B : L_k }
      { \Gamma \vdash \{m, n\} : F x:A.B }
      \rname{$F$-Intro}
    \end{mathpar}
    \vspace{-2em}
    \label{introduction}
    \Description{}
  \end{figure}

  \begin{figure}[h]
    \vspace{-0.8em}
    \caption{Select Elimination Rules}
    \begin{mathpar}
      \inferrule
      { \Gamma_1 \vdash m : \unit \\ 
        \Gamma_2 \vdash n : A \\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} }
      { \Gamma \vdash \letin{()}{m}{n} : A }
      \rname{Unit-Elim}

@@

      \inferrule
      { \Gamma_1 \vdash m : A \otimes B \\ 
        \Gamma_2, x \ltype A, y \ltype B \vdash n : C \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} }
      { \Gamma \vdash \letin{$[x, y]$}{$m$}{$n$} \vdash C }
      \rname{$\otimes$-Elim}

      \inferrule
      { \Gamma_1 \vdash m : F x:A.B \\
        \Gamma_2, x \utype A, y \ltype B \vdash n : C \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} }
      { \Gamma \vdash \letin{$\{x, y\}$}{$m$}{$n$} \vdash C }
      \rname{$F$-Elim}
    \end{mathpar}
    \vspace{-0.8em}
    \label{elimination}
    \Description{}
  \end{figure}

@@

  \subsection{Stateful Reasoning}
  We first postulate five state axioms. These axioms could be viewed as an interface to a trusted memory allocator.
  \begin{definition}
    State axioms.
    \begin{itemize}
      \item $l \mapsto_A m : \mathbb{N} \rightarrow (A : U_k) \rightarrow A \rightarrow L_k$ \\
      The $\mapsto$ is a linear type constructor. It fulfills a role similar to L3's \cite{l3} memory access capability, or Separation Logic's \cite{reynolds02} points-to assertion. Intuitively, it is a proof that a term $m$ of non-linear type $A$ is currently stored at address $l$.
      \item $\new : (A : U_k) \rightarrow (m : A) \rightarrow F l:\mathbb{N}.(l \mapsto_A m)$ \\
      $\new$ is a function that allocates a new memory store for term $m$ of non-linear type $A$. It results in an address $l$ paired with an assertion $l \mapsto_A m$.
      \item $\free : (A : U_k) \rightarrow F l:\mathbb{N}.F m:A.(l \mapsto_A m) \rightarrow \mathbf{1}$ \\
      $\free$ is a function that de-allocates the memory store at location $l$ which is currently in use. This can be seen in its input $F l:\mathbb{N}.F m:A.(l \mapsto_A m)$, where the $(l \mapsto_A m)$ component is a proof that location $l$ is currently allocated.
      \item $\get : (A : U_k) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \rightarrow F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$ \\
      $\get$ is a function that retrieves the term stored at address $l$ when given a proof $(l \mapsto_A m)$ that there is indeed a term $m$ currently stored there. The type of $\get$'s output is the rather complicated $F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$. This not only returns the term stored at $l$, but also a proof that the returned term is exactly the same as what $(l \mapsto_A m)$ claimed. The proof $(l \mapsto_A m)$ is returned unchanged as well.
      \item $\set : (A\ B : U_k) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \multimap (n : B) \multimap (l \mapsto_B n)$ \\
      $\set$ has the most complicated type of all the axioms presented here. Intuitively, if address $l$ is allocated and storing some term $m$ of type $A$, $\set$ may put term $n$ of type $B$ into the the store at address $l$, overwriting the original stored term. This is the so called "strong update". The consumption of $(l \mapsto_A m)$ and return of $(l \mapsto_B n)$ encode this update process.
    \end{itemize}
  \end{definition}
  Due to the fact that $(l \mapsto_A m)$ is a linear type, proofs of this type can not be duplicated and may only be consumed by other state axioms. This is the same non-sharing principle that empowers Separation Logic. Unlike Separation Logic, $(l \mapsto_A m)$ is a first-class object of CLC, meaning that it can be the result of computation (large eliminations), input to computations, stored in data structures etc.

  \subsection{Internalization of Effect Laws}
  The combination of linear types and dependent types within the state axioms allow the internalization of the so called "effect laws" for state, where these laws are provable theorems within the CLC framework.

  \begin{enumerate}
    \item Performing $\get$ twice on the same address will produce two terms that are provably equal.
    \item After using $\set$ to update the store at an address $l$ to term $m$, performing $\get$ again on the same address will retrieve $m$.
  \end{enumerate}

  Figure \ref{law1} is an excerpt taken from our implementation. It demonstrates the internalization of Law 1. The usage of implicit parameters allows us to omit much of the redundant arguments that could be inferred from the types of others. Here, $\new$ allocates a store for natural number 1, creating a proof $c$ of the form $(l \mapsto_\mathbb{N} 1)$. When $\get$ is applied to $c$ twice, the returned equality proofs $pf1$ and $pf2$ allow one to create proof $pf$ that the returned values $x$ and $y$ are equal. Finally, due to $c$ having linear type, it is required for it to be freed at the very end.

  \begin{figure}[h]
    \vspace{-1em}
    \caption{Law 1}
    \begin{lstlisting}
    let [ _, c ] := new _ 1 in
    let [ xeq, c ] := get _ _ _ c in
    let [ yeq, c ] := get _ _ _ c in
    let ( x, pf1 ) := xeq in
    let ( y, pf2 ) := yeq in
    let pf1 := Eq_sym _ _ _ pf1 in
    let pf : Eq _ x y := Eq_trans _ _ _ _ pf1 pf2 in
    free _ _ _ c.
    \end{lstlisting}
    \vspace{-1em}
    \label{law1}
    \Description{}
  \end{figure}

  Figure \ref{law2} demonstrates the internalization of Law 2. Again, implicit parameters fill in the annoying bureaucratic arguments when they can be inferred. Here, a store is allocated by $\new$ for natural number 1, creating a proof $c$ of the form $(l \mapsto_\mathbb{N} 1)$. After using $\set$ to update the store at address $l$ to 2 and then performing $\get$ on $l$, one can prove that the value retrieved by $\get$ is exactly equal to 2.

  \begin{figure}[h]
    \vspace{-1em}
    \caption{Law 2}
    \begin{lstlisting}
    let [ _, c ] := new _ 1 in
    let c := set _ _ _ _ c 2 in
    let [ zeq, c ] := get _ _ _ c in
    let ( z, pf1 ) := zeq in
    let pf : Eq _ 2 z := pf1 in
    free _ _ _ c.
    \end{lstlisting}
    \vspace{-1em}
    \label{law2}
    \Description{}
  \end{figure}

  Though we have not implemented program extraction for CLC yet, our preliminary experiments hint at the possibility of compiling the state axioms to constant time imperative operations after proof erasure. This can greatly increase the runtime efficiency of CLC over pure functional languages when used for programming. More examples are available in the first author's Github repository.

  \subsection{Mutable Data}
  Immutable data in functional programming languages are easy to reason about due their values staying constant over their lifespan. However, this convenience comes at the price of efficiency when attempting to perform "updates" on existing data. In order to update immutable data, an entirely new copy must be created with new entries substituted in for the original ones. So after update, there are two data structures in memory, the original one and the updated one. A garbage collector is necessary to reclaim memory allocated for the original copy.

  Due to the non-sharing property of linear types, it is possible to safely update linear data in-place without encountering synchronization issues. Garbage collection is no longer necessary as extra copies of data are not created. Languages such as ATS that implement mutable linear data have tiny runtime environments when compared to immutable garbage collection based ones such as OCaml or Haskell.

  \begin{figure}[h]
    \vspace{-1em}
    \caption{Linear Lists}
    \begin{lstlisting}
    Inductive list (A : Type) : Linear :=
    | nil : list A
    | cons : A -> list A -> list A.

    Fixpoint append (A : Type) : list A -o list A -o 
      (list A -o list A) -o list A :=
      fun ls1 ls2 k =>
        match ls1 with
        | nil => k ls2
        | (cons h t) as _cons_ => 
          append _ t ls2 (fun res => k (_cons_ h res))
        end.
    \end{lstlisting}
    \label{list}
    \vspace{-1em}
    \Description{}
  \end{figure}

  Figure \ref{list} presents a CPS append function for linear lists. In the \texttt{cons} case, the constructor is opened, exposing its data element \texttt{h} and tail list \texttt{t}. Instead of de-allocating the memory of \texttt{cons}, it is bound to the variable \texttt{_cons_} of linear type $A \multimap list\ A \multimap list\ A$. Intuitively, each \texttt{_cons_} must be applied exactly once to reconstruct a list using the memory of the original, thus mutating the original list in-place.

  \section{Future Work}
  The marriage of linear types and dependent types opens the door to many topics for research, both in theory and in its applications.

  On the theoretical side, we intend to investigate the semantic models of CLC, with the goal of explaining the seemingly natural cohesion between its rules. The impredicative nature of the \textit{Prop}-like universe $U_*$ hints at the possibility of extending term dependency to terms of linear types in a philosophically satisfying way. Linear types also appears to augment parametricity, which may produce stronger free theorems.

  On the application side, we intend to bolster our implementation with more features, making it a full-blown theorem prover/programming language. We aim to fully verify the typechecking algorithms and extensions employed by the implementation. The unification algorithm used to solve implicit parameters is extremely ad-hoc, as the problem of unification for linear types has not been thoroughly explored. We hope to improve this situation.

  \section{Conclusion}
  The Calculus of Linear Constructions is a lightweight extension of the Calculus of Constructions with linear types. As a type theory, it is a sound and effective framework for reasoning about resources. As a programing language, it allows one to write safe and efficient code, alleviating some of the issues faced by standard functional programming.
